---
title: "Lake Erie lmm-01 analysis"
author: "Michelle Berry"
date: "July 27, 2016"
output: html_document
---

#### This is an analysis to examine reads mapping to the lmm-01 microcystis phage genome from both the Lake Erie viromes and bacterial metagenomes collected in summer 2014. 


```{r global_options, echo = FALSE}

knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE
)

```

```{r}
library(ggplot2)
library(gtable)
library(gridExtra)
library(RColorBrewer)
library(cowplot)
library(dplyr)

theme_set(theme_grey())
mydir = "~/git_repos/mc-phage/"
```

First, read the coverage file to make a plot with x-axis: Total cumulated coverage, 
and y-axis: % of genome covered (across all erie viromes and metagenomes)
```{r}
# plotting theme
theme1 <- theme(
  axis.text = element_text(color = "black", size = 10), 
  axis.ticks = element_line(color = "black"), 
  legend.text = element_text(size = 8)
)


# read in data
lmm01_coverage <- read.csv("../data/lmm01-bam/All_lmm01_coverage.csv")

# plot
coverage_plot <- ggplot(data = lmm01_coverage, aes(x = coverage, y = length_covered, fill = sample)) + 
  geom_point(pch = 21, size = 5) + 
  xlab("Total cumulative coverage depth") +
  ylab("% of genome covered by Erie metagenome reads") + 
  ggtitle("Summary of coverage depth and breadth \n of lmm01 in Lake Erie samples") +
  theme1 

ggsave(filename = "../plots/coverage-plot.pdf", plot = coverage_plot, width = 6, height = 5)
```

Second, we read the detailed coverage file (listing all hits) to check the distribution 
of Id% and be able to do recruitment plots
```{r}
colors <- brewer.pal(4, "Set1")

# Read in detailed coverage data
recruit <- read.csv("../data/lmm01-bam/lmm01-recruitment.csv")
recruit$Genome <- "lmm01"

recruit$Sample <- factor(recruit$Sample, labels = c("ViromeJuly8", "ViromeAug4", "ViromeSept29", "ViromeOct20"))


# theme for recruitment distribution plots
theme_2 <- theme(
  axis.text = element_text(color = "black", size = 10),
  axis.ticks = element_line(color = "black"),
  legend.position = "none"
)

# Plot histogram
hist_1 <- ggplot(data = recruit, aes(x = Id)) + 
  geom_histogram(binwidth = 1, fill = "blue") + 
  theme_2 +
  ggtitle("lmm01 recruitment distribution across all samples")

ggsave(filename = "../plots/hist1.pdf", plot = hist_1, width = 6, height = 5)

hist_1

# Plot histograms separately for each sample 
hist_2 <- ggplot(data = recruit, aes(x = Id, fill = Sample)) + 
  geom_histogram(binwidth = 1) +
  facet_wrap(~Sample) + 
  theme_2 +
  ggtitle("lmm01 recruitment distribution per Sample")

ggsave(filename = "../plots/hist2.pdf", plot = hist_2, width = 6, height = 5)

hist_2
```

This plot shows there is very different coverage between samples
```{r fig.height=8, fig.width=6}

# Plot histograms separately for each sample adjusting the y axis
hist_3 <- ggplot(data = recruit, aes(x = Id, fill = Sample)) + 
  geom_histogram(binwidth = 1) +
  facet_grid(Sample~., scales = "free_y") + 
  theme_2 +
  ggtitle("lmm01 recruitment distribution per Sample \n adj. y-axis")

ggsave(filename = "../plots/hist3.pdf", plot = hist_3, width = 6, height = 8)

hist_3
```

Then we generate recruitment plots for each sample 
```{r}

# Max length of genome
max_length = max(recruit$Stop)

# Final matrix to store coverage
coverage <- matrix(data = NA, nrow = 0, ncol = 3)

# Change this value to modify the size of the sliding window, 
# larger sliding window (e.g. max_length/20) will be more smoothed, 
# smaller window (e.g. max_length/100) will be more detailed
window_size_half <- max_length/50  

# Can lower this value (e.g. max_length/100) if the coverage computation is too long
step <- max_length/500 
```

Max length is `r max_length`.     
Window size half is `r window_size_half`.    
Step is `r step`.     

```{r}
# Calculate coverage for each sample
for (samp in levels(recruit$Sample)) {
  
  # Filter dataframe to one sample
  recruit_sample <- filter(recruit, Sample == samp)
  
  # Genearte a matrix to hold coverage for each sample 
  coverage_sample <- matrix(0, max_length, 1)
  for (i in 1:nrow(recruit_sample)) {
    for (j in recruit_sample[i,2]:recruit_sample[i,3]) {
      coverage_sample[j,1] <- coverage_sample[j,1] + 1
    }
  }
  
  i = 1
  while (i < max_length) {
    total <- 2 * window_size_half 
    boundary_start <- i - window_size_half
    if (boundary_start < 0) {
      total <- total + boundary_start
      boundary_start <- 0
    }
    boundary_stop <- i + window_size_half
    if (boundary_stop > max_length) {
      total <- total - (boundary_stop - max_length)
      boundary_stop <- max_length
    }
    cover <- sum(coverage_sample[boundary_start:boundary_stop, 1])/total
    new_vec <- c(i, cover, samp)
    coverage <- rbind(coverage, new_vec)
    i <- i + step
  }
  
}

# Format coverage data.frame
coverage <- data.frame(coverage, row.names = NULL)
colnames(coverage) <- c("Coord", "Coverage", "Sample")
coverage$Coord <- as.numeric(as.character(coverage$Coord))
coverage$Coverage <- as.numeric(as.character(coverage$Coverage))

```


We have the coverage on sliding windows, let's do the plots 
```{r}  
  
# Start with the actual recruitment plot
plot_1 <- ggplot(data = recruit, aes(xmin = Start, xmax = Stop, ymin = Id, ymax = Id, color = Sample, name = Sample)) + 
  scale_y_continuous(limits = c(60, 101), breaks = c(60, 80, 100)) + 
  ylab("% Identity") + 
  xlab("Genomic position (bp)") + 
  scale_x_continuous(limits = c(0, max_length)) + 
  geom_rect(size = 0.7, alpha = 0.8) + 
  guides(fill = FALSE, col = guide_legend(override.aes = list(alpha = 1, size = 1 ))) + 
  ggtitle("lmm-01") +  
  theme(
    title = element_text(size = 8, lineheight = 0.8), 
    text = element_text(size = 8, lineheight = 0.8),
    legend.position = 'left',
    legend.justification = 'center', 
    legend.text = element_text(size = 6), legend.key.size = unit(0.5,"line"),
    axis.text = element_text(color = "black",size = 8),
    axis.ticks = element_line(color = "black")
  ) + 
  scale_color_manual(name = "Samples", values = colors)

ggsave("../plots/plot1.pdf")
plot_1

# Then the bottom coverage plot
ratio.display <- 50
ratio.values <- (max(coverage$Coord) - min(coverage$Coord))/(max(coverage$Coverage) -
  min(coverage$Coverage))

 plot_2 <- ggplot(data = coverage, aes(x = Coord, y = Coverage, group = Sample, color = Sample, name = Sample))  + 
  scale_x_continuous(limits = c(0, max_length)) + 
  scale_y_log10() + 
  ylab("Log10(coverage)") +  
  xlab("Genomic position (bp)") +
  geom_line(size = 0.4, alpha = 0.8) + 
  guides(
    fill = FALSE, 
    col = guide_legend(override.aes = list(alpha = 1, size = 3))
  ) + 
  theme(
    text = element_text(size = 8, lineheight = 0.8),  
    legend.position = 'left', 
    legend.justification = 'center',
    axis.text = element_text(color = "black", size = 8),
    axis.ticks = element_line(color = "black")
  ) + 
  scale_color_manual(values = colors) 

ggsave("../plots/plot2.pdf")
plot_2


# Then the read Id% distribution
plot_3 <- ggplot(data = recruit, aes(x = Id - 1)) + 
  geom_histogram(position = "identity", alpha = 0.6)  + 
  scale_x_continuous(limits = c(60, 101)) + 
  coord_flip()  +  
  ylab("Counts") + 
  guides(fill = FALSE, col = guide_legend(override.aes = list(alpha = 1, size = 3), ncol = 1)) + 
  ggtitle("") + 
  theme(
    title = element_text(size = 8,lineheight = .8), 
    text = element_text(size = 8,lineheight = .8),
    legend.position = 'right',
    legend.justification = 'center', 
    axis.title.y = element_blank(),
    axis.text = element_text(color = "black", size = 8),
    axis.ticks = element_line(color = "black")
  )

ggsave("../plots/plot3.pdf")
plot_3
```

Then we put all of these together as nicely as we can
```{r fig.width=12, fig.height=8}

  
gp <- ggplotGrob(plot_1)
gp2 <- ggplotGrob(plot_2)
gp3 <- ggplotGrob(plot_3)
maxWidth = grid::unit.pmax(gp$widths[2:5], gp2$widths[2:5])
gp$widths[2:5] <- as.list(maxWidth)
gp2$widths[2:5] <- as.list(maxWidth)
pdf(file = "../plots/lmm01-combined-plot.pdf", width = 13, height = 6, useDingbats = F)
final_plot <- grid.arrange(gp, gp3, gp2, nrow = 2, ncol = 2, widths = c(8,2))
dev.off()

ggdraw() +
  draw_plot(final_plot, x = 0.04, y = 0, width = .96, height = .96)

```